---
title: "Introduction to SLGP Package"
author: "Athénaïs Gautier"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette:  >
  %\VignetteIndexEntry{Introduction to SLGP Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette serves as a startup guide to SLGP modeling, providing a practical introduction to the implementation of Spatial Logistic Gaussian Processes (SLGPs). 

# Dataset

We illustrate the model's capabilities using the `quakes` dataset shipped with base R [@r_core_team]: 1000 seismic events of body-wave magnitude above 4.0 recorded since 1964 in the Fiji-Tonga region, originating from the Harvard PRIM-H project. Each event carries its epicentre (`lat`, `long`), its hypocentre `depth` (40-680 km), its magnitude `mag`, and the number of `stations` that detected it.

For this vignette, we focus on modeling the distribution of hypocentre depths (`depth`) along an east-west transect, indexed by longitude (`long`). The region hosts two subduction systems, and the resulting Wadati-Benioff zones [@isacks_seismology_1968; @frohlich_deep_2006] make the conditional distribution of `depth` deform across the transect - from shallow-unimodal near the trench, to bimodal where shallow and deep seismicity coexist, to deep-dominated. Together with the naturally clustered epicentres (no replicates, wildly uneven coverage), this makes the catalogue an ideal stress test for SLGP modeling.

```{r loadQuakes, warning=FALSE, message=FALSE}
data("quakes")
library(tidyr)
library(dplyr)

# Bin data together for visualisation purpose
df <- quakes %>%
  mutate(long_bin = cut(long, breaks = seq(165, 190, by = 2.5), include.lowest = FALSE)) %>%
  group_by(long_bin) %>%
  mutate(long_bin = paste0(long_bin, "\nn=", n()))%>%
  ungroup()%>%
  mutate(long_bin = factor(long_bin, 
                           levels = sort(unique(long_bin), decreasing = FALSE))) %>%
  data.frame()

range_response <- c(40, 680) # Can use range(df$depth), or user defined range as we do here
range_x <- c(165, 190) # Can use range(df$long), or user defined range as we do here
```

We represent the data to visualise the relationship between `depth` and `long`.

```{r figureQuakes, fig.cap = "A visual representation of the dependency of hypocentre depth on longitude in the Fiji-Tonga `quakes` catalogue.", fig.fullwidth=TRUE, fig.height=4, fig.width=10, fig.align='center',fig.pos="H"}
library(ggplot2)
library(ggpubr)
library(viridis)
# Scatterplot: long vs depth
scatter_plot <- ggplot(df, aes(x = long, y = depth)) +
  geom_point(alpha = 0.5, color = "grey", pch=20) +
  labs(y = "Hypocentral depth (km)",
       x = "Longitude (°)",
       title = "Depth measured at various longitudes") +
  theme_bw()+
  coord_cartesian(xlim=range_x,
                  ylim=range_response)

# Histogram: Distribution of depth by 'long' bin
hist_plot <- ggplot(df, aes(x = depth)) +
  geom_histogram(mapping=aes(y=after_stat(density)),
                 position = "identity", breaks = seq(40, 680, 40),
                 fill="darkgrey", col="grey50", lwd=0.2, alpha=0.7) +
  geom_rug(sides = "b", color = "navy", alpha = 0.5)+
  facet_wrap(~ long_bin, scales = "free_y", nrow=2) +
  labs(x = "Longitude (°)",
       y = "Probability density", 
       title = "Histogram of 'depth' by 'longitude' group") +
  theme_bw()+
  coord_cartesian(xlim=range_response,
                  ylim=c(0, 0.015))
ggarrange(scatter_plot, hist_plot, ncol = 2, nrow = 1,
          widths = c(0.3, 0.7))
```

The two planes of seismicity are visible to the naked eye: a shallow band near the trenches and a deep band, down to nearly 700 km, along the subducting slab. Between the two, both regimes coexist and the empirical distributions turn bimodal. This dataset provides a demanding test case for SLGP modeling, offering a compact, one-dimensional covariate space, heterogeneously distributed data, and shifting distributional shapes with a well-understood geophysical ground truth.

# SLGP model specification

To model the distributional changes observed in the earthquake catalogue, we now introduce the Spatial Logistic Gaussian Process (SLGP) model. SLGPs provide a flexible non-parametric framework for modeling spatially dependent probability densities. By transforming a Gaussian Process (GP) through exponentiation and normalization, SLGPs ensure positivity and integration to one, making them well-suited for density estimation.

## Maximum a posteriori estimate

For a quick approximation, MAP estimation provides a point estimate, offering a balance between computational efficiency and the depth of inference. It is the fastest estimation scheme we propose, however MAP does not facilitate uncertainty quantification because it yields a non-probabilistic estimate of the underlying density field, focusing instead on identifying the mode of the posterior distribution. 

```{r SLGPfitting1}
library(SLGP)
modelMAP <- slgp(depth~long, # Use a formula to specify predictors VS response
                 # Can use depth~. for all variables,
                 # Or 'depth' ~ long + var2 + var3 for more variables
                 data=df,
                 method="MAP", # MAP estimation scheme
                 basisFunctionsUsed = "RFF",
                 interpolateBasisFun="WNN", # Will Accelerate inference
                 hyperparams = list(lengthscale=c(0.15, 0.15), 
                                    # Applied to normalised data
                                    # So 0.15 is 15% of the range of values
                                    sigma2=1), 
                 # Will be re-selected with sigmaEstimationMethod
                 sigmaEstimationMethod = "heuristic", 
                 # Set to heuristic for numerical stability                 
                 predictorsLower= c(range_x[1]),
                 predictorsUpper= c(range_x[2]),
                 responseRange= range_response,
                 opts_BasisFun = list(nFreq=200,
                                      MatParam=5/2),
                 seed=1)
summary(modelMAP)
```

We can represent the conditional densities. We first use the standard `plot()` method for `SLGP` objects.

```{r SLGPplottingPrior1, fig.cap = "Conditional depth densities across longitude under the MAP estimate of the SLGP.", fig.fullwidth=TRUE, fig.height=5, fig.width=10, fig.align='center', fig.pos="H"}
plot(modelMAP, panels = TRUE)
```

We can also use the `predict()` method for `SLGP` objects and make more advanced visuals.



```{r SLGPplotting2, fig.cap = "Predictive probability density of 'depth' at 'long', seen over slices.", fig.fullwidth=TRUE, fig.height=5, fig.width=10, fig.align='center',fig.pos="H"}
library(viridis)
dfGrid <- data.frame(expand.grid(seq(range_x[1], range_x[2], 1), 
                                 seq(range_response[1], range_response[2],, 101)))
colnames(dfGrid) <- c("long", "depth")
pred <- predict(modelMAP, newdata= dfGrid)

scale_factor <- 300
ggplot()  +
  labs(x = "Hypocentral depth (km)",
       y = "Longitude (°)")+
  theme_bw()+
  geom_ribbon(data=pred,
              mapping=aes(x=depth, ymax=scale_factor*pdf_1+long, 
                          ymin=long, group=-long, fill=long),
              col="grey", alpha=0.9)+
  geom_point(data=df,
             mapping=aes(x = depth, y = long), alpha = 0.5, 
             pch=20, color = "grey")+
  scale_fill_viridis(option = "plasma",
                     guide = guide_colorbar(title = "Indexing variable: Longitude",
                                            barheight = unit(2, units = "mm"),
                                            barwidth = unit(55, units = "mm"),
                                            title.position = 'top',
                                            label.position = "bottom",
                                            title.hjust = 0.5))+
  theme(legend.position = "bottom")+
  coord_flip()
```


The figure reveals how the SLGP captures the structure in the estimated conditional density of `depth` given `long`. It effectively adapts as the shallow mode near the trench gives way to the deep Wadati-Benioff mode along the transect. Notably, the model tracks the bimodal transition region, assigning probability mass to both regimes where they coexist.

### Comparing the estimation to data

```{r SLGPplottingMAP, fig.cap = "Binned 'depth' histograms by 'long' (width = 1) VS SLGP MAP estimators at bins centers", fig.fullwidth=TRUE, fig.height=4, fig.width=10, fig.align='center',fig.pos="H"}

selected_values <- c(167, 180, 185)
gap <- 0.5
df_filtered <- df %>%
  mutate(interval=findInterval(long, c(0, 
                                       selected_values[1]-gap, 
                                       selected_values[1]+gap, 
                                       selected_values[2]-gap, 
                                       selected_values[2]+gap, 
                                       selected_values[3]-gap, 
                                       selected_values[3]+gap)))%>%
  filter(interval %in% c(2, 4, 6))%>%
  group_by(interval)%>%
  mutate(category = paste0("long close to ", c("", selected_values[1],
                                               "", selected_values[2],
                                               "", selected_values[3])[interval], 
                           "\nn=", n()))
names <- sort(unique(df_filtered$category))
dfGrid <- data.frame(expand.grid(selected_values, 
                                 seq(range_response[1], range_response[2],, 101)))
colnames(dfGrid) <- c("long", "depth")
predMAP <- predict(modelMAP, newdata = dfGrid)
colnames(predMAP) <- c("long", "depth", "MAP estimator")
predMAP <- predMAP%>%
  pivot_longer(-c("long", "depth"))
predMAP$category <-ifelse(predMAP$long==selected_values[1], names[1],
                          ifelse(predMAP$long==selected_values[2], names[2], names[3]))

ggplot(mapping=aes(x = depth)) +
  geom_histogram(df_filtered,
                 mapping=aes(y=after_stat(density)),
                 position = "identity", breaks = seq(40, 680, 20),
                 fill="darkgrey", col="grey50", lwd=0.2, alpha=0.7) +
  geom_rug(data=df_filtered, sides = "b", color = "navy", alpha = 0.5)+
  geom_line(data=predMAP, mapping=aes(y=value, group=name), 
            color = "black", lwd=0.1, alpha=0.5)+
  geom_line(data=predMAP, mapping=aes(y=value, group=name, col=name), lwd=1.1)+
  facet_wrap(~ category, scales = "free_y", nrow=1) +
  labs(x = "Hypocentral depth (km)",
       y = "Probability density",
       title = "Binned 'depth' histograms by 'long' (width = 1) VS SLGP-MAP estimators at bins centers") +
  theme_bw()+
  theme(legend.position="bottom",
        legend.direction = "horizontal",
        legend.title = element_blank())+
  coord_cartesian(xlim=range_response,
                  ylim=c(0, 0.02))
```

This other figure directly compares the SLGP MAP estimate to histograms of `depth` at selected `long` values, illustrating how well the model aligns with the observed data. Despite the absence of replicates, the SLGP estimate follows the empirical distribution, adapting to changes in shape - and in the number of modes - across the transect.

## Laplace approximation estimate

By integrating the MAP approach with Laplace approximation, we refine our estimation strategy by approximating the posterior distribution with a multivariate Gaussian. This method strikes a balance between the full Bayesian inference of MCMC and the computational efficiency of MAP estimation. By leveraging both the gradient and Hessian of the posterior, it captures essential curvature information, providing a more informed approximation of the posterior landscape.

We can train a SLGP from scratch using the argument 'method="Laplace"', or re-training a model. 


```{r SLGPfitting2}
modelLaplace <- update(modelMAP, newdata = df, method="Laplace")
summary(modelLaplace)
```

```{r SLGPfitting22, eval=FALSE}
# Or equivalent, more explicit in the re-using of the elements
# From the SLGP prior
modelLaplace <- slgp(medv~age, 
                     data=df,
                     method="Laplace", #Maximum a posteriori estimation scheme
                     basisFunctionsUsed = "RFF",
                     interpolateBasisFun="WNN", # Accelerate inference
                     hyperparams = modelMAP@hyperparams, 
                     sigmaEstimationMethod = "none",# Already selected in the prior
                     predictorsLower= c(range_x[1]),
                     predictorsUpper= c(range_x[2]),
                     responseRange= range_response,
                     opts_BasisFun = modelMAP@opts_BasisFun,
                     BasisFunParam = modelMAP@BasisFunParam,
                     seed=1)
```


Unlike MAP estimation, which provides only a point estimate, the Laplace approximation allows us to visualize uncertainty.


```{r SLGPLaplaceplo0, fig.cap = "Predictive probability density of 'depth' at 'long', as predicted by a SLGP with Laplace approximation", fig.fullwidth=TRUE, fig.height=5, fig.width=10, fig.align='center', fig.pos="H"}
plot( modelLaplace, draw = c("mean", 1:10), panels = TRUE)
```


To highlight the uncertainty quantification, we compare multiple posterior draws from the approximation against histogram data, emphasizing how the estimated conditional density fluctuates. The figure below illustrates this by overlaying samples from the posterior approximation with binned data, showcasing the range of possible densities rather than focusing solely on the MAP estimate.

```{r SLGPLaplaceplot, fig.cap = "Predictive probability density (and draws from a Laplace approximation) of 'depth' at 'long', seen over 3 slices.", fig.fullwidth=TRUE, fig.height=4, fig.width=8, fig.align='center',fig.pos="H"}

dfGrid <- data.frame(expand.grid(seq(3), 
                                 seq(range_response[1], range_response[2],, 101)))
colnames(dfGrid) <- c("ID", "depth")
dfGrid$long <- selected_values[dfGrid$ID]
pred <- predict(modelLaplace, newdata = dfGrid)
pred$meanpdf <- rowMeans(pred[, -c(1:3)])

library(tidyr)
# Filter the data: keep values within ±5 of the selected ones
df_filtered <- df %>%
  mutate(interval=findInterval(long, c(0, 
                                       selected_values[1]-gap, 
                                       selected_values[1]+gap, 
                                       selected_values[2]-gap, 
                                       selected_values[2]+gap, 
                                       selected_values[3]-gap, 
                                       selected_values[3]+gap)))%>%
  filter(interval %in% c(2, 4, 6))%>%
  group_by(interval)%>%
  mutate(category = paste0("long close to ", c("", selected_values[1],
                                               "", selected_values[2],
                                               "", selected_values[3])[interval], 
                           "\nn=", n()))
names <- sort(unique(df_filtered$category))
pred$category <- names[pred$ID]

set.seed(1)
selected_cols <- sample(seq(1000), size=10, replace=FALSE)
df_plot <- pred %>%
  dplyr::select(c("long", "depth", "category", 
                  paste0("pdf_", selected_cols)))%>%
  pivot_longer(-c("long", "depth", "category"))


ggplot(mapping=aes(x = depth)) +
  geom_histogram(df_filtered,
                 mapping=aes(y=after_stat(density)),
                 position = "identity", breaks = seq(40, 680, 20),
                 fill="darkgrey", col="grey50", lwd=0.2, alpha=0.7) +
  geom_rug(data=df_filtered, sides = "b", color = "navy", alpha = 0.5)+
  geom_line(data=df_plot, mapping=aes(y=value, group=name), 
            color = "black", lwd=0.1, alpha=0.5)+
  geom_line(data=pred, mapping=aes(y=meanpdf, group=category), color = "red")+
  facet_wrap(~ category, scales = "free_y", nrow=1) +
  labs(x = "Hypocentral depth (km)",
       y = "Probability density",
       title = "Binned 'depth' histograms by 'long' (width = 1) VS SLGP-Laplace estimators at bins centers") +
  theme_bw()+
  coord_cartesian(xlim=range_response,
                  ylim=c(0, 0.02))

```


## MCMC estimate

This method allows us to explore the posterior distribution by drawing samples from it. It enables precise, exact inference of the posterior distribution of the underlying density field knowing the data. The main drawback of this approach being its higher computational cost

```{r SLGPfitting3, eval=FALSE}
modelMCMC <- slgp(medv~age, # Use a formula to specify predictors VS response
                  # Can use medv~. for all variables,
                  # Or medv ~ age + var2 + var3 for more variables
                  data=df,
                  method="MCMC", #MCMC
                  basisFunctionsUsed = "RFF",
                  interpolateBasisFun="WNN", # Accelerate inference
                  hyperparams = list(lengthscale=c(0.15, 0.15), 
                                     # Applied to normalised data
                                     # So 0.15 is 15% of the range of values
                                     sigma2=1), 
                  # Will be re-selected with sigmaEstimationMethod
                  sigmaEstimationMethod = "heuristic", # Set to heuristic for numerical stability                 
                  predictorsLower= c(range_x[1]),
                  predictorsUpper= c(range_x[2]),
                  responseRange= range_response,
                  opts_BasisFun = list(nFreq=100,
                                       MatParam=5/2),
                  opts = list(stan_chains=2, stan_iter=1000))
```


```{r SLGPfitting3b, eval=FALSE}
modelMCMC <- update(modelMAP, 
                    newdata = df, 
                    method="MCMC", opts=list(stan_iter=100, stan_chains=2))
summary(modelMCMC)
```

# By-products of the estimation

One of the advantages of the SLGP framework is that it predicts entire probability density functions (PDFs) over space. This opens the door to a wide range of nonlinear inferences on the estimated field. In particular, we can compute and visualize functionals of the predicted densities, such as moments (mean, variance, skewness, etc.) or quantiles.

In our current implementation, we provide predictions for both centered and uncentered moments of the estimated PDFs at each location. These derived quantities can themselves be interpreted as spatial fields and are thus informative summaries of the underlying random process.

Importantly, when using probabilistic inference schemes like Laplace approximation or MCMC, the uncertainty in the SLGP predictions naturally propagates to these functionals. As a result, we can also quantify uncertainty on the functionals. For example, we compute credible intervals for the mean or variance field. This is illustrated in the upcoming figure, where we use the MCMC-trained SLGP model to compute moments.


```{r SLGPMCMCplotMoments, fig.cap = "Simultaneous prediction of the fields moments (and associated uncertainty) using a SLGP model", fig.fullwidth=TRUE, fig.height=4, fig.width=10, fig.align='center',fig.pos="H"}
dfX <- data.frame(long=seq(range_x[1], range_x[2], 1))

predMean <- predict(modelLaplace, type= "moments", newdata = dfX, 
                    power=c(1),
                    centered=FALSE) # Uncentered moments
# For the mean
predVar <- predict(modelLaplace, type= "moments", newdata = dfX, 
                   power=c(2, 3, 4),
                   centered=TRUE) # Centered moments
# For the variance, Kurtosis and Skewness

pred <- rbind(predMean, predVar)
pred <- pred %>%
  pivot_longer(-c("long", "power"))%>%
  mutate(value=ifelse(power==2, sqrt(value), value))%>% # Define std
  pivot_wider(values_from = value,
              names_from = power)%>%
  mutate(`3`=`3`/`2`^2,
         `4`=`4`/`2`^4)%>% # Kurtosis and Skewness
  pivot_longer(-c("long", "name"), names_to = "power")%>%
  data.frame()

pred$power <- factor(c("Expected value", 
                       "Standard deviation",
                       "Skewness","Kurtosis")[as.numeric(pred$power)],
                     levels=c("Expected value", "Standard deviation",
                              "Skewness", "Kurtosis"))
df_plot <- pred %>%
  group_by(long, power)%>%
  summarise(q10 = quantile(value, probs=c(0.1)),
            q50 = quantile(value, probs=c(0.5)),
            q90 = quantile(value, probs=c(0.9)),
            mean = mean(value), .groups="keep")%>%
  ungroup() # summarise uncertainty


ggplot(df_plot, mapping=aes(x = long, group=power)) +
  geom_ribbon(mapping = aes(ymin=q10, ymax=q90),
              alpha = 0.25, lty=2, col="black", fill="cornflowerblue")+
  geom_line(mapping=aes(y=q50))+
  facet_wrap(.~power,
             scales = "free", nrow=1)+
  labs(x =  "Longitude (°)", 
       y = "Moment value") +
  theme_bw()+
  coord_cartesian(xlim=range_x)

```

We also support the joint prediction of quantiles at arbitrary levels. Because quantiles are derived directly from the estimated densities, they are guaranteed to be consistent and non-crossing - a property that cannot always be ensured in standard quantile regression. This makes them reliable tools for summarizing distributional shape (e.g., asymmetry, spread) at each location.


```{r SLGPMCMCplotQuantiles2, fig.cap = "Simultaneous quantile prediction (and associated uncertainty) using a SLGP model, with estimation performed by MCMC", fig.fullwidth=TRUE, fig.height=5, fig.width=10, fig.align='center',fig.pos="H", warning=FALSE, message=FALSE}
probsL <-  c(10, 25, 50, 75, 90)/100

# SLGP prediction
pred <- predict(modelLaplace, type = "quantiles",
                newdata = dfX,
                probs = probsL)

df_plot <- pred %>%
  pivot_longer(-c("long", "probs"))%>%
  group_by(long, probs)%>%
  summarise(q10 = quantile(value, probs=c(0.1)),
            q50 = quantile(value, probs=c(0.5)),
            q90 = quantile(value, probs=c(0.9)),
            mean = mean(value), .groups="keep")%>%
  ungroup()%>%
  mutate(probs=factor(paste0("Quantile: ", 100*probs, "%"),
                      levels=paste0("Quantile: ", 100*probsL, "%")))


plot1 <-ggplot(df_plot, mapping=aes(x = long)) +
  geom_ribbon(mapping = aes(ymin=q10, ymax=q90, 
                            col=probs, fill=probs, group=probs),
              alpha = 0.25, lty=3)+
  geom_line(mapping=aes(y=q50, lty="SLGP", 
                        col=probs,group=probs), lwd=0.75)+
  labs(x = "Longitude [°E]",
       y = "Hypocentre depth [km]",
       col = "Quantile levels",
       fill = "Quantile levels",
       lty = "Quantile estimation method") +
  theme_bw() +
  coord_cartesian(xlim = range_x, ylim = range_response) 

plot2 <- ggplot(df, mapping = aes(x = long)) +
  geom_histogram(col = "navy", fill = "grey", alpha = 0.3,
                 breaks = seq(range_x[1], range_x[2], 1)) +
  theme_minimal() +
  labs(x = NULL, y = "Sample\ncount")


p_with_marginal <- ggarrange(
  plot2, plot1,
  ncol = 1, heights = c(1, 3),  # adjust height ratio
  align = "v"
)

# Display the plot
print(p_with_marginal)
```

# References

